\chapter{Development and Implementation}\label{cha:dev}

\section{Development Strategy}\label{sec:dev_strat}

The main goal of this study was to create a tool that can automatically classify \gls{DNA} sequences using \gls{ML}/\gls{DL} models, followed by its integration into \textit{ProPythia}. The other key objective was to integrate automated machine learning classifiers into the \textit{OmniumAI} software platforms.

However, it is crucial to understand the steps and the technologies necessary to build such tools. These tools were built with \textit{Python}, which is a high-level interpreted general-purpose programming language that supports vast and extensive external libraries that are constantly evolving. \textit{PyCharm Code Editor} was used to combine \textit{Python} code development, while also improving usability and user experience. Additionally, the \textit{Anaconda} software application was employed to facilitate package management and distribution. Then, for building the entire classification pipeline, the \textit{PyTorch} framework was used as it is one of the most popular free open source and powerful \gls{DL} frameworks.

In an \gls{ML} project, after selecting the dataset, the first step is the data processing task since, in most cases, the data is not ready to be fed into a model. The sequences need first to be converted into a numerical representation, and how this step is executed has a substantial influence on the model's ultimate performance. 
As mentioned in Section~\ref{sec:workflow}, data processing can be subdivided into feature extraction and feature selection. The first is the process of transforming raw data into suitable modelling features, and the second consists of filtering irrelevant features, keeping the essential ones. If working with a \gls{DL} model, the user may not need to perform this step since these kind of models can extract the features from the training data on their own.

In this study, descriptors were chosen to be the features for shallow \gls{ML} classification models, which is a manual feature extraction process. For \gls{DL} models, they only need the sequences as input, but they still need to be in a numerical format. Encoders were utilized to address this problem, since they can create numerical representations of a string of letters.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\input{Chapters/6.1_setting_up_data}

\input{Chapters/6.2_classifiers}
