%!TEX root = ../template.tex
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% acknowledgements.tex
%% NOVA thesis document file
%%
%% Text with acknowledgements
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\acknowledgements

To begin with, I would like to thank my supervisor, Prof. Miguel Rocha for giving me this opportunity and for all the support he provided whenever I needed it. Then, I would also like to thank Ana Marta Sequeira for the continuous support and guidance throughout the whole dissertation, as well as the patience for all my questions. This project was only possible with their help, and I'm forever grateful for it.

I would also like to thank the colleagues from OmniumAI, namely Fernando Cruz, João Capela Ribeiro and Miguel Barros for the huge help regarding the new technologies I had to learn and the challenges I had to face.

Then, I want to thank my parents for providing me all the opportunities and everything I have ever asked for. Thank you for all the support and encouragement throughout my whole life, but especially this last year.

I also want to thank my friends who were present in my life during these last 5 years of university, especially Hugo, Tiago, and Duarte. Thank you all for the amazing moments we have spent together and for being the best friends I have ever had.

Lastly, I would like to thank all the family, friends, and colleagues I did not mention here, but who were present in my life during these last few years. Thank you all for the support and encouragement you have provided me.